---
title: "data_preprocessing"
author: "Jerry Xu"
date: "11 October 2018"
output:
  # pdf_document: default
  html_document: default
---

```{r, include=TRUE}

# library(tidyverse)
library(dplyr)
library(ggplot2)
source("./functions/cv_knn.R")
source("./functions/cv_da.R")
source("./functions/cv_rpart.R")
```


In the previous markdown file we explored the dataset, relabelling the variables as NA as appropriate according to the data dictionary. 

Visualisations and other exploratory analysis was conducted mostly on a "cleaned" dataset with all missing variables dropped by row and column, which is not a reliable method of data preprocessing. 

Hence, we want to preprocess the dataset properly by first assessing which variables are necessary and handling the missing data/outliers methodically.

We utilise the AHS dataset that has the appropriate variable outcomes relabelled as NA with nothing dropped. 
 
```{r}

feature_variables = c("BMISC", "AGEC", "PHDKGWBC", "EXLWTBC", "SF2SA1QN", "INCDEC", "HSUGBC", "FATT1", "SUGART1", 
                       "PREVAT1", "PROVAT1", "FATPER1", "LAPER1", "ALAPER1", "CHOPER1", "SUGPER1", "SATPER1", "TRANPER1", 
                       "MONOPER1", "POLYPER1", "ADTOTSE", "SEX", "SMKSTAT", "SYSTOL", "FASTSTAD", "HDLCHREB", "LDLNTR", "LDLRESB", "B3T1")

# BMR, SLPTIME

response_variables = c("CHOLNTR", "HDLCHREB", "DIABBC", "HCHOLBC", "HYPBC", "CVDMEDST")

all_variables = c(feature_variables, response_variables)

nutm_orig = read.csv("../output/nutmstat_factors_and_NAs.csv")
nutm = nutm_orig[all_variables]
```

We'll need to reprocess some of the variables as factors. 

```{r}
categoricalList <- c()

categoricalList[ 1 ] <- FALSE #  BMISC 
categoricalList[ 2 ] <- FALSE #  AGEC 

categoricalList[ 3 ] <- FALSE #  PHDKGWBC 

categoricalList[ 4 ] <- FALSE #  EXLWTBC 

categoricalList[ 5 ] <- TRUE #  SF2SA1QN 
categoricalList[ 6 ] <- TRUE #  INCDEC 

categoricalList[ 7 ] <- TRUE #  HSUGBC 

categoricalList[ 8 ] <- FALSE #  FATT1 

categoricalList[ 9 ] <- FALSE #  SUGART1 

categoricalList[ 10 ] <- FALSE #  PREVAT1 
categoricalList[ 11 ] <- FALSE #  PROVAT1 

categoricalList[ 12 ] <- FALSE #  FATPER1 
categoricalList[ 13 ] <- FALSE #  LAPER1 
categoricalList[ 14 ] <- FALSE #  ALAPER1 
categoricalList[ 15 ] <- FALSE #  CHOPER1 
categoricalList[ 16 ] <- FALSE #  SUGPER1 

categoricalList[ 17 ] <- FALSE #  SATPER1 
categoricalList[ 18 ] <- FALSE #  TRANPER1 

categoricalList[ 19 ] <- FALSE #  MONOPER1 
categoricalList[ 20 ] <- FALSE #  POLYPER1 
categoricalList[ 21 ] <- FALSE #  ADTOTSE 

categoricalList[ 22 ] <- TRUE #  SEX 

categoricalList[ 23 ] <- TRUE #  SMKSTAT 
categoricalList[ 24 ] <- FALSE #  SYSTOL 

categoricalList[ 25 ] <- TRUE #  FASTSTAD 

categoricalList[ 26 ] <- TRUE #  HDLCHREB 
categoricalList[ 27 ] <- TRUE #  LDLNTR 
categoricalList[ 28 ] <- TRUE #  LDLRESB 

categoricalList[ 29 ] <- TRUE #  B3T1

categoricalList[ 30 ] <- TRUE #  CHOLNTR 
categoricalList[ 31 ] <- TRUE #  HDLCHREB 
categoricalList[ 32 ] <- TRUE #  DIABBC 
categoricalList[ 33 ] <- TRUE #  HCHOLBC 
categoricalList[ 34 ] <- TRUE #  HYPBC 
categoricalList[ 35 ] <- TRUE #  CVDMEDST 

# "CHOLNTR", "HDLCHREB", "DIABBC", "HCHOLBC", "HYPBC", "CVDMEDST"

for (i in 1:length(categoricalList)) {
  if (categoricalList[ i ]) {
      nutm[,i] <- as.factor(nutm[  ,i])
  }
}

head(nutm)
```

## Missing Values

First we check for missing values in this particular subset

```{r}
dim(nutm)
idx_len = length(all_variables)

percent_missing_values <- data.frame(matrix(ncol=idx_len, nrow=3))
colnames(percent_missing_values) <- all_variables

for (i in 1:length(nutm)){
  
  percent_missing_values[1, i] = sum(is.na(nutm[, i]))
  percent_missing_values[2, i] = dim(nutm)[1] - sum(is.na(nutm[, i]))
  percent_missing_values[3, i] = percent_missing_values[1, i]/dim(nutm)[1]
}


```

```{r}

round_2dp <- function(x){
  round(x, 2)
} 

percent_missing_values <- data.frame(lapply(percent_missing_values, round_2dp))

rownames(percent_missing_values) <- c("missing", "not_missing", "percent_missing")

# percent_missing_values[3, ]
column_idx <- c(1:idx_len)

#plot(column_idx, percent_missing_values[3, ], type="h", 
#     main="% missing values for each column", ylab="% missing values", xlab = "column_index")

```

```{r}

pmv_transpose <- data.frame(t(percent_missing_values))
pmv_transpose

ggplot(pmv_transpose, aes(x=column_idx, y=(percent_missing*100), fill="purple")) + geom_bar(stat="identity") +
  xlab("variables") + ylab("% missing") + ggtitle("% Values Missing in a Variable") +
  guides(fill=FALSE) 

```

Looking at this, we may have to omit some variables depending on which response variables we are analying.


## Predicting Diabetes Mellitus

We use to model the "DIABBC" variable. It contains 3 outcomes:

```{r}

table(nutm["DIABBC"])

```

\begin{tabular}{|c|c|}
\hline
Outcome & Interpretation    \\
\hline
1       & Currently has DB  \\
3       & Had DB previously \\
5       & Does not have DB  \\
\hline
\end{tabular}

Judging by the distribution of the different responses, we may have to drop 3 entirely as the definition of such an outcome is unclear. We do not know how long ago they stopped, although a retrospective study of their health improvements could be an interesting further piece of work.

Given the disparity between the binary outcomes, a Poisson model here makes sense. 

Now we need to select our feature variables. 

```{r}
diabetes_response = c("DIABBC")
diabetes_variables = c("BMISC", "AGEC", "EXLWTBC", 
                       # "SF2SA1QN", "INCDEC",
                       # "HSUGBC", 
                       "FATT1", "SUGART1", 
                       "FATPER1", "LAPER1", "ALAPER1", "CHOPER1", 
                       "SATPER1", "TRANPER1", "MONOPER1", "POLYPER1", 
                       "ADTOTSE"
                       # "SEX"
                       # "SMKSTAT", "SYSTOL", "FASTSTAD", 
                       # "HDLCHREB", "LDLNTR", "LDLRESB", 
                       #"B3T1"
                       )

diabetes = nutm[c(diabetes_response, diabetes_variables)]
diabetes = diabetes[diabetes$DIABBC != 3, ]

table(diabetes$DIABBC)

diabetes <- mutate(diabetes, DIABBC = ifelse(DIABBC == 5, "0", "1"))
diabetes$DIABBC <- as.factor(diabetes$DIABBC)
head(diabetes)
```

```{r}
sum(is.na(nutm[diabetes_response]))

diabetes_no_na <- diabetes[complete.cases(diabetes), ]

dim(diabetes_no_na)
table(diabetes_no_na$DIABBC)
```

It seems like dropping all rows with any NA values does not change the binary outcomes of the response by alot. This is good. 

```{r}

ggplot(diabetes_no_na, aes(x=BMISC)) + stat_density(aes(fill=DIABBC)) +
  labs(title ="Diabetes", x = "BMI", y = "density") + 
  scale_fill_discrete(labels=c("Neg", "Pos"))

ggplot(diabetes_no_na, aes(x=SUGART1)) + stat_density(aes(fill=DIABBC)) +
  labs(title ="Diabetes", x = "Sugar", y = "density") + 
  scale_fill_discrete(labels=c("Neg", "Pos"))

```


Now we can try to fit a KNN to test the code. 

```{r}
neighbours=3
V=10

X = data.frame(scale(diabetes_no_na[, 2:15]))
y = diabetes_no_na$DIABBC
head(X)
cv_knn(X, y, k=neighbours, V=10, seed=1)

res.rpart <- rpart(m_y_nz ~ m_x_nz
, data=df_cart_no_zeroes)
rpart.plot(res.rpart,type=1,extra=1, main="CART for Diabetes, non_zeroes")
```

```{r}
X <- as.matrix(diabetes_no_na[, 2:15])
y <- diabetes_no_na$DIABBC

diabetes_no_na_cart <- data.frame(y, X)

res_rpart <- rpart(y ~ X, data=diabetes_no_na_cart)
res_rpart
rpart.plot(res_rpart,type=1,extra=1, main="CART for Diabetes")
```

```{r}

X = data.frame(scale(diabetes_no_na[, 2:15]))
y = diabetes_no_na$DIABBC

cv_da(X = X, y = y, V = 10, method = "lda", seed = 1)
cv_da(X = X, y = y, V = 10, method = "qda", seed = 1)

```

```{r}

X = data.frame(scale(diabetes_no_na[, 2:15]))
y = diabetes_no_na$DIABBC

mutate(diabetes_no_na, DIABBC = ifelse(DIABBC == "0", 0, 1))

logistic_regression = glm(y~., family=binomial, data=X)
summary(logistic_regression)

```

