---
title: "Data_preprocessing"
author: "Jerry Xu"
date: "11 October 2018"
output:
  # pdf_document: default
  html_document: default
---

Warning: The glm functions take a while to run. They have been commented out to save time when running the whole file

```{r, include=TRUE}

# library(tidyverse)
library(tidyr)
library(dplyr)
library(broom) # for tidy()
library(ggplot2)

```


In the previous markdown file we explored the dataset, relabelling the variables as NA as appropriate according to the data dictionary. 

Visualisations and other exploratory analysis was conducted mostly on a "cleaned" dataset with all missing variables dropped by row and column, which is not a reliable method of data preprocessing. 

Hence, we want to preprocess the dataset properly by first assessing which variables are necessary and handling the missing data/outliers methodically.

We utilise the AHS dataset that has the appropriate variable outcomes relabelled as NA with nothing dropped. 
 
```{r}

feature_variables = c("BMISC", "AGEC", "PHDKGWBC", "EXLWTBC", "SF2SA1QN", "INCDEC", "HSUGBC", "FATT1", "SUGART1", 
                       "PREVAT1", "PROVAT1", "FATPER1", "LAPER1", "ALAPER1", "CHOPER1", "SUGPER1", "SATPER1", "TRANPER1", 
                       "MONOPER1", "POLYPER1", "ADTOTSE", "SEX", "SMKSTAT", "SYSTOL", "FASTSTAD", "HDLCHREB", "LDLNTR", "LDLRESB", "B3T1")

# BMR, SLPTIME

response_variables = c("CHOLNTR", "HDLCHREB", "DIABBC", "HCHOLBC", "HYPBC", "CVDMEDST")

all_variables = c(feature_variables, response_variables)

nutm_orig = read.csv("../output/nutmstat_factors_and_NAs.csv")
nutm = nutm_orig[all_variables]

```

```{r}

table(nutm$CHOLNTR) # Total Cholesterol Status
table(nutm$HDLCHREB) # HDL Cholesterol Range
table(nutm$LDLRESB) # LDL Cholesterol Range
table(nutm$DIABBC) # Diabetes Mellitus status
table(nutm$HCHOLBC) # High Cholesterol Status 
table(nutm$HYPBC) # Hypertensive Disease status
table(nutm$CVDMEDST) # Dyslipidemia status (1-3 are similar)

```

We'll need to reprocess some of the variables as factors. 

```{r}
categoricalList <- c()

categoricalList[ 1 ] <- FALSE #  BMISC 
categoricalList[ 2 ] <- FALSE #  AGEC 

categoricalList[ 3 ] <- FALSE #  PHDKGWBC 

categoricalList[ 4 ] <- FALSE #  EXLWTBC 

categoricalList[ 5 ] <- TRUE #  SF2SA1QN 
categoricalList[ 6 ] <- TRUE #  INCDEC 

categoricalList[ 7 ] <- TRUE #  HSUGBC 

categoricalList[ 8 ] <- FALSE #  FATT1 

categoricalList[ 9 ] <- FALSE #  SUGART1 

categoricalList[ 10 ] <- FALSE #  PREVAT1 
categoricalList[ 11 ] <- FALSE #  PROVAT1 

categoricalList[ 12 ] <- FALSE #  FATPER1 
categoricalList[ 13 ] <- FALSE #  LAPER1 
categoricalList[ 14 ] <- FALSE #  ALAPER1 
categoricalList[ 15 ] <- FALSE #  CHOPER1 
categoricalList[ 16 ] <- FALSE #  SUGPER1 

categoricalList[ 17 ] <- FALSE #  SATPER1 
categoricalList[ 18 ] <- FALSE #  TRANPER1 

categoricalList[ 19 ] <- FALSE #  MONOPER1 
categoricalList[ 20 ] <- FALSE #  POLYPER1 
categoricalList[ 21 ] <- FALSE #  ADTOTSE 

categoricalList[ 22 ] <- TRUE #  SEX 

categoricalList[ 23 ] <- TRUE #  SMKSTAT 
categoricalList[ 24 ] <- FALSE #  SYSTOL 

categoricalList[ 25 ] <- TRUE #  FASTSTAD 

categoricalList[ 26 ] <- TRUE #  HDLCHREB 
categoricalList[ 27 ] <- TRUE #  LDLNTR 
categoricalList[ 28 ] <- TRUE #  LDLRESB 

categoricalList[ 29 ] <- FALSE #  B3T1

categoricalList[ 30 ] <- TRUE #  CHOLNTR 
categoricalList[ 31 ] <- TRUE #  HDLCHREB 
categoricalList[ 32 ] <- TRUE #  DIABBC 
categoricalList[ 33 ] <- TRUE #  HCHOLBC 
categoricalList[ 34 ] <- TRUE #  HYPBC 
categoricalList[ 35 ] <- TRUE #  CVDMEDST 

# "CHOLNTR", "HDLCHREB", "DIABBC", "HCHOLBC", "HYPBC", "CVDMEDST"

for (i in 1:length(categoricalList)) {
  if (categoricalList[ i ]) {
      nutm[,i] <- as.factor(nutm[ ,i])
  }
}

head(nutm)
```

## Missing Values

First we check for missing values in this particular subset

```{r}
dim(nutm)
idx_len = length(all_variables)

percent_missing_values <- data.frame(matrix(ncol=idx_len, nrow=3))
colnames(percent_missing_values) <- all_variables

for (i in 1:length(nutm)){
  
  percent_missing_values[1, i] = sum(is.na(nutm[, i]))
  percent_missing_values[2, i] = dim(nutm)[1] - sum(is.na(nutm[, i]))
  percent_missing_values[3, i] = percent_missing_values[1, i]/dim(nutm)[1]
}


```

```{r}

round_2dp <- function(x){
  round(x, 2)
} 

percent_missing_values <- data.frame(lapply(percent_missing_values, round_2dp))

rownames(percent_missing_values) <- c("missing", "not_missing", "percent_missing")

# percent_missing_values[3, ]
column_idx <- c(1:idx_len)

#plot(column_idx, percent_missing_values[3, ], type="h", 
#     main="% missing values for each column", ylab="% missing values", xlab = "column_index")

```

```{r}

pmv_transpose <- data.frame(t(percent_missing_values))
pmv_transpose

ggplot(pmv_transpose, aes(x=column_idx, y=(percent_missing*100), fill="purple")) + geom_bar(stat="identity") +
  xlab("variables") + ylab("% missing") + ggtitle("% Values Missing in a Variable") +
  guides(fill=FALSE) 

```

Looking at this, we may have to omit some variables depending on which response variables we are analying.
