---
title: "Cholesterol"
author: "Jerry Xu"
date: "16 October 2018"
output: html_document
---

## Predicting Abnormal Levels of Cholesterol

Warning: The glm functions take a while to run. They have been commented out to save time when running the whole file

```{r, include=TRUE}

# library(tidyverse)
library(tidyr)
library(dplyr)
library(broom) # for tidy()
library(ggplot2)

# library(rsample)      # data splitting - not available for the most recent version of R
library(randomForestSRC) # basic RF package
library(ranger)

source("./functions/cv_knn.R")
source("./functions/cv_da.R")
source("./functions/cv_rpart.R")
source("./functions/cv_glm.R")
source("./functions/cv_glm_backward.R")

```

## Standardised Preprocessing

```{r}

feature_variables = c("BMISC", "AGEC", "PHDKGWBC", "EXLWTBC", "SF2SA1QN", "INCDEC", "HSUGBC", "FATT1", "SUGART1", 
                       "PREVAT1", "PROVAT1", "FATPER1", "LAPER1", "ALAPER1", "CHOPER1", "SUGPER1", "SATPER1", "TRANPER1", 
                       "MONOPER1", "POLYPER1", "ADTOTSE", "SEX", "SMKSTAT", "SYSTOL", "FASTSTAD", "HDLCHREB", "LDLNTR", "LDLRESB", "B3T1")

# BMR, SLPTIME

response_variables = c("CHOLNTR", "HDLCHREB", "DIABBC", "HCHOLBC", "HYPBC", "CVDMEDST")

all_variables = c(feature_variables, response_variables)

nutm_orig = read.csv("../output/nutmstat_factors_and_NAs.csv")
nutm = nutm_orig[all_variables]

```

```{r}

table(nutm$CHOLNTR) # Total Cholesterol Status
table(nutm$HDLCHREB) # HDL Cholesterol Range
table(nutm$LDLRESB) # LDL Cholesterol Range
table(nutm$DIABBC) # Diabetes Mellitus status
table(nutm$HCHOLBC) # High Cholesterol Status 
table(nutm$HYPBC) # Hypertensive Disease status
table(nutm$CVDMEDST) # Dyslipidemia status (1-3 are similar)

```

We'll need to reprocess some of the variables as factors. 

```{r}
categoricalList <- c()

categoricalList[ 1 ] <- FALSE #  BMISC 
categoricalList[ 2 ] <- FALSE #  AGEC 

categoricalList[ 3 ] <- FALSE #  PHDKGWBC 

categoricalList[ 4 ] <- FALSE #  EXLWTBC 

categoricalList[ 5 ] <- TRUE #  SF2SA1QN 
categoricalList[ 6 ] <- TRUE #  INCDEC 

categoricalList[ 7 ] <- TRUE #  HSUGBC 

categoricalList[ 8 ] <- FALSE #  FATT1 

categoricalList[ 9 ] <- FALSE #  SUGART1 

categoricalList[ 10 ] <- FALSE #  PREVAT1 
categoricalList[ 11 ] <- FALSE #  PROVAT1 

categoricalList[ 12 ] <- FALSE #  FATPER1 
categoricalList[ 13 ] <- FALSE #  LAPER1 
categoricalList[ 14 ] <- FALSE #  ALAPER1 
categoricalList[ 15 ] <- FALSE #  CHOPER1 
categoricalList[ 16 ] <- FALSE #  SUGPER1 

categoricalList[ 17 ] <- FALSE #  SATPER1 
categoricalList[ 18 ] <- FALSE #  TRANPER1 

categoricalList[ 19 ] <- FALSE #  MONOPER1 
categoricalList[ 20 ] <- FALSE #  POLYPER1 
categoricalList[ 21 ] <- FALSE #  ADTOTSE 

categoricalList[ 22 ] <- TRUE #  SEX 

categoricalList[ 23 ] <- TRUE #  SMKSTAT 
categoricalList[ 24 ] <- FALSE #  SYSTOL 

categoricalList[ 25 ] <- TRUE #  FASTSTAD 

categoricalList[ 26 ] <- TRUE #  HDLCHREB 
categoricalList[ 27 ] <- TRUE #  LDLNTR 
categoricalList[ 28 ] <- TRUE #  LDLRESB 

categoricalList[ 29 ] <- FALSE #  B3T1

categoricalList[ 30 ] <- TRUE #  CHOLNTR 
categoricalList[ 31 ] <- TRUE #  HDLCHREB 
categoricalList[ 32 ] <- TRUE #  DIABBC 
categoricalList[ 33 ] <- TRUE #  HCHOLBC 
categoricalList[ 34 ] <- TRUE #  HYPBC 
categoricalList[ 35 ] <- TRUE #  CVDMEDST 

# "CHOLNTR", "HDLCHREB", "DIABBC", "HCHOLBC", "HYPBC", "CVDMEDST"

for (i in 1:length(categoricalList)) {
  if (categoricalList[ i ]) {
      nutm[,i] <- as.factor(nutm[ ,i])
  }
}

head(nutm)
```

#### Data Preprocessing Stage

Let's firstly check how much data we have.

```{r}

sum(is.na(nutm["CHOLNTR"]))
table(nutm["CHOLNTR"])

```

```{r}

cholesterol_response = c("CHOLNTR")
cholesterol_variables = c("BMISC", "AGEC", "EXLWTBC", 
                       # "SF2SA1QN", "INCDEC",
                       # "HSUGBC", 
                       "FATT1", "SUGART1", 
                       "FATPER1", "LAPER1", "ALAPER1", "CHOPER1", 
                       "SATPER1", "TRANPER1", "MONOPER1", "POLYPER1", 
                       # "ADTOTSE",
                       # "SEX"
                       # "SMKSTAT", 
                       "SYSTOL", 
                       # "FASTSTAD", 
                       # "HDLCHREB", 
                       # "LDLNTR", 
                       # "LDLRESB", 
                       "B3T1"
                       )

cholesterol = nutm[c(cholesterol_response, cholesterol_variables)]

table(cholesterol$CHOLNTR)

# cholesterol <- mutate(cholesterol, CHOLNTR = ifelse(CHOLNTR == 1, "0", "1"))
cholesterol$CHOLNTR <- as.factor(cholesterol$CHOLNTR)
# cholesterol$HDLCHREB <- as.numeric(cholesterol$HDLCHREB)
cholesterol$LDLNTR <- as.numeric(cholesterol$LDLNTR)
# cholesterol$LDLRESB <- as.numeric(cholesterol$LDLRESB)
head(cholesterol)

```

```{r}

cholesterol_no_na <- cholesterol[complete.cases(cholesterol), ]
dim(cholesterol_no_na)

table(cholesterol_no_na$CHOLNTR)

1855/(1069+1855)
```

Setting variables:

```{r}

df_len = dim(cholesterol_no_na)[2]
df_len

neighbours = 3
V = 10

```


```{r}

X = data.frame(scale(cholesterol_no_na[, 2:df_len]))
y = cholesterol_no_na$CHOLNTR
cv_knn(X, y, k=neighbours, V=10, seed=1)


k_neighbours = seq(1,29,by=2)
cv_errors = c()
for (i in 1:length(k_neighbours)) {
  cv_errors[i] = cv_knn(X, as.factor(y),k=k_neighbours[i],V,seed=1)
}
plot(k_neighbours, cv_errors, type="l", xlab="value of k", ylab="cross-validation error", cex.lab=1.5,
    main="CV errors for K-nearest neighbours", cex.main=2, ylim=c(0.0,0.20))

```

```{r}

X <- as.matrix(cholesterol_no_na[, 2:df_len])
y <- cholesterol_no_na$CHOLNTR
cholesterol_no_na_cart <- data.frame(y, X)

cv_rpart(X,y,V)

res_rpart <- rpart(y ~ X
, data=cholesterol_no_na)
rpart.plot(res_rpart,type=1,extra=1, main="CART for Cholesterol, non_zeroes")
res_rpart

cv_rpart(X, y, V, seed=1)

```

### LDA and QDA

```{r}

X = data.frame(scale(cholesterol_no_na[, 2:15]))
y = cholesterol_no_na$CHOLNTR

cv_da(X = X, y = y, V = 10, method = "lda", seed = 1)
cv_da(X = X, y = y, V = 10, method = "qda", seed = 1)

```

#### GLMs

```{r}

X = data.frame(cholesterol_no_na[, 2:df_len])
y = cholesterol_no_na$CHOLNTR

mutate(cholesterol_no_na, CHOLNTR = ifelse(CHOLNTR == "0", 0, 1))

logistic_regression = glm(y~., family=binomial, data=X)
summary(logistic_regression)

```

Now let's try to conduct some cross-validation on GLMs. 

```{r, eval=FALSE, include=FALSE}

cv_glm(X, y, V = 10)
cv_glm_backward(X, y, V = 10, pen = log(df_samples))
cv_glm_backward(X, y, V = 10, pen = 2)

```


#### Random Forests

```{r}

X = data.frame(cholesterol_no_na[, 2:df_len])
y = cholesterol_no_na$CHOLNTR

cholesterol_no_na_cart <- data.frame(y, X)

train <- sample_frac(cholesterol_no_na_cart, 0.7)
train_idx <-as.numeric(rownames(train))
test <- cholesterol_no_na_cart[-train_idx,]
head(train)
head(test)

```

```{r}

chol_train <- ranger(
  formula = y ~ .,
  data    = train,
  importance = "impurity"
  # xtest   = test$y,
  # ytest   = test[, 2:df_len]
)

chol_train$variable.importance
chol_train$prediction.error

print("Predict using the test set")
chol_test <- predict(chol_train, data = test)

table(test$y, chol_test$predictions)

chol_train$variable.importance %>% 
  tidy() %>%
  dplyr::arrange(desc(x)) %>%
  dplyr::top_n(25) %>%
  ggplot(aes(reorder(names, x), x)) +
  geom_col() +
  coord_flip() +
  ggtitle("Top 25 important variables")

```

__Tuning the Hyperparameters__

```{r}
hyper_grid_2 <- expand.grid(
  mtry       = seq(1, 14, by = 2),
  node_size  = seq(3, 9, by = 2),
  sampe_size = c(.55, .632, .70, .80),
  OOB_RMSE  = 0
)

hyper_grid_2

# perform grid search
for(i in 1:nrow(hyper_grid_2)) {
  
  # train model
  model <- ranger(
    formula         = y ~ ., 
    data            = train, 
    importance      = "impurity",
    num.trees       = 500,
    mtry            = hyper_grid_2$mtry[i],
    min.node.size   = hyper_grid_2$node_size[i],
    sample.fraction = hyper_grid_2$sampe_size[i],
    seed            = 123
  )
  
  # add OOB error to grid
  hyper_grid_2$OOB_RMSE[i] <- sqrt(model$prediction.error)
}

hyper_grid_2 %>% 
  dplyr::arrange(OOB_RMSE) %>%
  head(10)
```
