---
title: "Project"
author: "Ian Astalosh"
date: "04/10/2018"
output: html_document
---

```{r Read data, echo=FALSE}
dat_orig <- read.csv("nutmstatData2018.csv",header=TRUE)

dat <-  dat_orig
colNames <- colnames(dat_orig)
n = nrow(dat)

numerics = c()
categoricals = c()
indexes = c()
indexes_cat = c()
```

```{r data cleaning, echo=FALSE}
##PUT THIS AT THE END

BMISC_exclude <- c(0,98,99)
dat$BMISC[dat$BMISC%in%BMISC_exclude] <- NA

##AGEC no cleaning required, there is no one aged 99+

dat$SMSBC = as.factor(dat$SMSBC)

dat$COBBC = as.factor(dat$COBBC)

dat$FEMLSBC[dat$FEMLSBC == 9] = NA
dat$FEMLSBC = as.factor(dat$FEMLSBC)

##is it worth keeping what type of NA it is?
PHDKGWBC_exclude <- c(997,998,999)
dat$PHDKGWBC[dat$PHDKGWBC%in%PHDKGWBC_exclude] <- NA

##worth keeping the NA?
PHDCMHBC_exclude <- c(0,998,999)
dat$PHDCMHBC[dat$PHDCMHBC%in%PHDCMHBC_exclude] <- NA

##For EXLWTBC, EXLWMBC, EXLWVBC, I'm assuming 9996 are NA's so will be removed as such.
EXLWTBC_exclude <- c(9996,9999)
dat$EXLWTBC[dat$EXLWTBC%in%EXLWTBC_exclude] <- NA

EXLWMBC_exclude <- c(9996,9999)
dat$EXLWMBC[dat$EXLWMBC%in%EXLWMBC_exclude] <- NA

EXLWVBC_exclude <- c(9996,9999)
dat$EXLWVBC[dat$EXLWVBC%in%EXLWVBC_exclude] <- NA

PHDCMWBC_exclude <- c(0,998,999)
dat$PHDCMWBC[dat$PHDCMWBC%in%PHDCMWBC_exclude] <- NA

BMR_exclude <- 99998
dat$BMR[dat$BMR%in%BMR_exclude] <- NA

EIBMR1_exclude <- 998
dat$EIBMR1[dat$EIBMR1%in%EIBMR1_exclude] <- NA

SF2SA1QN_exclude <- c(0,99)
dat$SF2SA1QN[dat$SF2SA1QN%in%SF2SA1QN_exclude] <- NA
dat$SF2SA1QN = as.factor(dat$SF2SA1QN)

INCDEC_exclude <- c(0,98,99)
dat$INCDEC[dat$INCDEC%in%INCDEC_exclude] <- NA
dat$INCDEC = as.factor(dat$INCDEC)

dat$DIABBC = as.factor(dat$DIABBC)

dat$HCHOLBC = as.factor(dat$HCHOLBC)

dat$HSUGBC = as.factor(dat$HSUGBC)

dat$HYPBC = as.factor(dat$HYPBC)

ADTOTSE_exclude <- c(0,9996,9999)
dat$ADTOTSE[dat$ADTOTSE%in%ADTOTSE_exclude] <- NA

BDYMSQ04_exclude <- c(0,6)
dat$BDYMSQ04[dat$BDYMSQ04%in%BDYMSQ04_exclude] <- NA
dat$BDYMSQ04 = as.factor(dat$BDYMSQ04)

DIASTOL_exclude <- c(0,998,999)
dat$DIASTOL[dat$DIASTOL%in%DIASTOL_exclude] <- NA

#are not known, not used valid NA values? if so, add in extra things to exclude
DIETQ12_exclude <- c(0,6) ##may either need to include or exclude 5 or 6
dat$DIETQ12[dat$DIETQ12%in%DIETQ12_exclude] <- NA
dat$DIETQ12 = as.factor(dat$DIETQ12)

DIETQ14_exclude <- c(0,6) ##may either need to include or exclude 5 or 6
dat$DIETQ14[dat$DIETQ14%in%DIETQ14_exclude] <- NA
dat$DIETQ14 = as.factor(dat$DIETQ14)

DIETQ5_exclude <- 0
dat$DIETQ5[dat$DIETQ5%in%DIETQ5_exclude] <- NA
dat$DIETQ5 = as.factor(dat$DIETQ5)

DIETQ8_exclude <- 0
dat$DIETQ8[dat$DIETQ8%in%DIETQ8_exclude] <- NA
dat$DIETQ8 = as.factor(dat$DIETQ8)

DIETRDI_exclude <- c(0,3)
dat$DIETRDI[dat$DIETRDI%in%DIETRDI_exclude] <- NA
dat$DIETRDI = as.factor(dat$DIETRDI)

SABDYMS_exclude <- c(0,8,9)
dat$SABDYMS[dat$SABDYMS%in%SABDYMS_exclude] <- NA
dat$SABDYMS = as.factor(dat$SABDYMS)

dat$SEX = as.factor(dat$SEX)

SLPTIME_exclude <- c(0,9998,9999)
dat$SLPTIME[dat$SLPTIME%in%SLPTIME_exclude] <- NA

SMKSTAT_exclude <- 0
dat$SMKSTAT[dat$SMKSTAT%in%SMKSTAT_exclude] <- NA
dat$SMKSTAT = as.factor(dat$SMKSTAT)

SYSTOL_exclude <- c(0,998,999)
dat$SYSTOL[dat$SYSTOL%in%SYSTOL_exclude] <- NA

ALTNTR_exclude <- c(0,8)
dat$ALTNTR[dat$ALTNTR%in%ALTNTR_exclude] <- NA
dat$ALTNTR = as.factor(dat$ALTNTR)

ALTRESB_exclude <- c(97,98)
dat$ALTRESB[dat$ALTRESB%in%ALTRESB_exclude] <- NA
dat$ALTRESB = as.factor(dat$ALTRESB)

APOBNTR_exclude <- c(0,8)
dat$APOBNTR[dat$APOBNTR%in%APOBNTR_exclude] <- NA
dat$APOBNTR = as.factor(dat$APOBNTR)

APOBRESB_exclude <- c(97,98)
dat$APOBRESB[dat$APOBRESB%in%APOBRESB_exclude] <- NA
dat$APOBRESB = as.factor(dat$APOBRESB)

B12RESB_exclude <- c(97,98)
dat$B12RESB[dat$B12RESB%in%B12RESB_exclude] <- NA
dat$B12RESB = as.factor(dat$B12RESB)

BIORESPC_exclude <- 0
dat$BIORESPC[dat$BIORESPC%in%BIORESPC_exclude] <- NA
dat$BIORESPC = as.factor(dat$BIORESPC)

CHOLNTR_exclude <- c(0,8)
dat$CHOLNTR[dat$CHOLNTR%in%CHOLNTR_exclude] <- NA
dat$CHOLNTR = as.factor(dat$CHOLNTR)

CHOLRESB_exclude <- c(97,98)
dat$CHOLRESB[dat$CHOLRESB%in%CHOLRESB_exclude] <- NA
dat$CHOLRESB = as.factor(dat$CHOLRESB)

CVDMEDST_exclude <- c(0,8)
dat$CVDMEDST[dat$CVDMEDST%in%CVDMEDST_exclude] <- NA
dat$CVDMEDST = as.factor(dat$CVDMEDST)

DIAHBRSK_exclude <- c(0,8)
dat$DIAHBRSK[dat$DIAHBRSK%in%DIAHBRSK_exclude] <- NA
dat$DIAHBRSK = as.factor(dat$DIAHBRSK)

FASTSTAD_exclude <- 0
dat$FASTSTAD[dat$FASTSTAD%in%FASTSTAD_exclude] <- NA
dat$FASTSTAD = as.factor(dat$FASTSTAD)

FOLATREB_exclude <- c(97,98)
dat$FOLATREB[dat$FOLATREB%in%FOLATREB_exclude] <- NA
dat$FOLATREB = as.factor(dat$FOLATREB)

GGTNTR_exclude <- 8
dat$GGTNTR[dat$GGTNTR%in%GGTNTR_exclude] <- NA
dat$GGTNTR = as.factor(dat$GGTNTR)

GGTRESB_exclude <- c(97,98)
dat$GGTRESB[dat$GGTRESB%in%GGTRESB_exclude] <- NA
dat$GGTRESB = as.factor(dat$GGTRESB)

GLUCFPD_exclude <- c(0,8)
dat$GLUCFPD[dat$GLUCFPD%in%GLUCFPD_exclude] <- NA
dat$GLUCFPD = as.factor(dat$GLUCFPD)

GLUCFREB_exclude <- c(97,98)
dat$GLUCFREB[dat$GLUCFREB%in%GLUCFREB_exclude] <- NA
dat$GLUCFREB = as.factor(dat$GLUCFREB)

HBA1PREB_exclude <- c(7,8)
dat$HBA1PREB[dat$HBA1PREB%in%HBA1PREB_exclude] <- NA
dat$HBA1PREB = as.factor(dat$HBA1PREB)

HDLCHREB_exclude <- c(7,8)
dat$HDLCHREB[dat$HDLCHREB%in%HDLCHREB_exclude] <- NA
dat$HDLCHREB = as.factor(dat$HDLCHREB)

LDLNTR_exclude <- c(0,8)
dat$LDLNTR[dat$LDLNTR%in%LDLNTR_exclude] <- NA
dat$LDLNTR = as.factor(dat$LDLNTR)

LDLRESB_exclude <- c(97,98)
dat$LDLRESB[dat$LDLRESB%in%LDLRESB_exclude] <- NA
dat$LDLRESB = as.factor(dat$LDLRESB)

TRIGNTR_exclude <- c(0,8)
dat$TRIGNTR[dat$TRIGNTR%in%TRIGNTR_exclude] <- NA
dat$TRIGNTR = as.factor(dat$TRIGNTR)

TRIGRESB_exclude <- c(0,97, 98)
dat$TRIGRESB[dat$TRIGRESB%in%TRIGRESB_exclude] <- NA
dat$TRIGRESB = as.factor(dat$TRIGRESB)

SMKDAILY_exclude <- c(0)
dat$SMKDAILY[dat$SMKDAILY%in%SMKDAILY_exclude] <- NA
dat$SMKDAILY = as.factor(dat$SMKDAILY)
```

```{r, echo=FALSE}
for (i in 1:length(colNames)){
 if(is.factor(dat[,i]) == FALSE){
   numerics = c(numerics, colNames[i])
   indexes = c(indexes, i) } 
  else {
    categoricals = c(categoricals, colNames[i])
    indexes_cat = c(indexes_cat, i) }
}

numeric_only = dat[,indexes]
categorical_only = dat[,indexes_cat]
```

```{r}
length(dat$CVDMEDST)
table(dat$CVDMEDST)
sum(is.na(dat$CVDMEDST))
```


```{r}
library(tidyverse)
##PREDICTORS OF INTEREST
predictors = select(dat, BMISC, AGEC, PHDKGWBC, EXLWTBC, SF2SA1QN, INCDEC, HSUGBC, FATT1, SUGART1, PREVAT1, PROVAT1, FATPER1, LAPER1, ALAPER1, CHOPER1, SATPER1, SUGPER1, TRANPER1, MONOPER1, POLYPER1, ADTOTSE, SEX, SMKSTAT, SYSTOL, FASTSTAD, HDLCHREB, LDLNTR, LDLRESB, B3T1)

##RESPONSES OF INTEREST
responses = select(dat, CHOLNTR, HDLCHREB, DIABBC, HCHOLBC, HYPBC, CVDMEDST)

predNames = colnames(predictors)
respNames = colnames(responses)

##PREDICTORS 
numerics=c()
indexes=c()
categoricals = c()
indexes_cat=c()

for (i in 1:length(predNames)){
 if(is.factor(predictors[,i]) == FALSE){
   numerics = c(numerics, predNames[i])
   indexes = c(indexes, i) } 
  else {
    categoricals = c(categoricals, predNames[i])
    indexes_cat = c(indexes_cat, i) }
}

numeric_only_preds = predictors[,indexes]
categorical_only_preds = predictors[,indexes_cat]

##PREDICTOR CATEGORICAL DENSITY PLOTS
for (i in 1:dim(numeric_only_preds)[2]){
  plot(density(na.omit(numeric_only_preds[,i])))
}

numeric_only_preds = predictors[,indexes]
categorical_only_preds = predictors[,indexes_cat]

for (i in 1:dim(numeric_only_preds)[2]){
 boxplot(na.omit(numeric_only_preds[,i]))
}
```
```{r}
#PERCENTAGE MISSING IN OUR VARIABLES OF INTEREST
total = cbind(responses, predictors)
image(is.na(t(total)))

perc_miss_tot = apply(is.na(total), 2, mean)
missing_data_tot = data.frame(Names = names(perc_miss_tot), Percent_Missing = perc_miss_tot)

missing_data_tot

```

There is extremely high missingness in total cholsterol, HDL cholsterol and Dyslipidemia. It doesn't really make sense to impute the outcome vector, so create a new matrix that removes NA's in the outcomes:

```{r}

#Probably a faster way of doing this, but this systematically removes the rows containing missing values in each column
rem1 = total[!is.na(total$CHOLNTR),]
rem2 = total[!is.na(total$HDLCHREB),]
rem3 = total[!is.na(total$CVDMEDST),]

dim(rem3) 
image(is.na(t(rem3)))

for (i in 1:dim(rem3)[2]){
 boxplot(na.omit(rem3[,i]))
}

perc_miss_new = 100*apply(is.na(rem3), 2, mean)
missing_data_new = data.frame(Var = names(perc_miss_new), Percent_Missing = perc_miss_new)
missing_data_new
```
The variable containing the most missingness is income per household (INCDEC) at like 5%. From here, we can either remove the existing rows containing NA's, potentially remove INCDEC (if we really wanted to, since socio economic status and incdec ) or impute the missingness from here. 

```{r}
omitted_rem3 = na.omit(rem3)
dim(omitted_rem3)
100*dim(omitted_rem3)[1]/dim(rem3)[1]
```
Removing all NA's results in a loss of 13% of the entries (which I don't think is terrible, we still have 2718 observations)

If we remove INDEC (which is probably related to socio-economic factor, although they only have 36% correlation) and then omit NA's then:

```{r}
noinc = rem3[, -12]
new = na.omit(noinc)
dim(new)
100*dim(new)[1]/dim(rem3)[1]
```
ie. only lose 92%. Maybe worth bringing up with group to see what variables are valid and not valid to be dropped. 

```{r}
rem3 = na.omit(rem3)
rem3names = colnames(rem3)

numerics=c()
indexes=c()
categoricals = c()
indexes_cat=c()

for (i in 1:length(rem3names)){
 if(is.factor(rem3[,i]) == FALSE){
   numerics = c(numerics, rem3names[i])
   indexes = c(indexes, i) } 
  else {
    categoricals = c(categoricals, rem3names[i])
    indexes_cat = c(indexes_cat, i) }
}

rem3_num = rem3[,indexes]
rem3_cat = rem3[,indexes_cat]

for (i in 1:dim(rem3_num)[2]){
 boxplot(log(na.omit(rem3_num[,i])))
}

for (i in 1:dim(rem3_num)[2]){
 plot(density(na.omit(rem3_num[,i])))
}
```

Basic exploratory analysis of types of fat
```{r}
plotdf = data.frame(`Carbohydrate` = rem3$CHOPER1, `Total Sugars` = rem3$SUGPER1, `Total Fat` = rem3$FATPER1, `Linoleic Acid` = rem3$LAPER1, `Alpha Linolenic Acid` = rem3$ALAPER1, `Saturated Fat + Fatty Acids` = rem3$SATPER1, `Trans-Fatty Acids` = rem3$TRANPER1, `Monounsaturated Fat` = rem3$MONOPER1, `Polyunsaturated Fat` = rem3$POLYPER1)

melted = melt(plotdf)
colnames(melted) = c("Type", "Value")

ggplot(melted, aes(x = Type, y = Value)) + geom_boxplot() + ggtitle("Energy Sources (Percentage of Energy)") + theme(plot.title = element_text(hjust = 0.5, size=12), axis.text.x = element_text(angle=90, hjust=1))
```

Boxplots of the fats only
```{r}
fatsonly = data.frame(`Linoleic Acid` = rem3$LAPER1, `Alpha Linolenic Acid` = rem3$ALAPER1, `Saturated Fat + Fatty Acids` = rem3$SATPER1, `Trans-Fatty Acids` = rem3$TRANPER1, `Monounsaturated Fat` = rem3$MONOPER1, `Polyunsaturated Fat` = rem3$POLYPER1)

melted = melt(fatsonly)
colnames(melted) = c("Type", "Value")

ggplot(melted, aes(x = Type, y = Value)) + geom_boxplot() + ggtitle("Types of Fat (Percentage of Energy))") + theme(plot.title = element_text(hjust = 0.5, size=12), axis.text.x = element_text(angle=90, hjust=1))
```

```{r}
library(corrplot)
cormat = cor(fatsonly)
corrplot(cormat, method="color",  addCoef.col = "black", type = "upper", tl.col="black", tl.srt=45)


```

```{r}
agetrenddf = data.frame(Age = rem3$AGEC, Chol = as.numeric(rem3$CHOLNTR))

for (j in 1:length(agetrenddf$Chol)){
  if (agetrenddf$Chol[j] == 1){
    agetrenddf$Chol[j] = "Normal Total Cholesterol"
  }
  if (agetrenddf$Chol[j] == 2){
    agetrenddf$Chol[j] = "Abnormal Total Cholesterol"
  }
}

cats = c()

for (i in 1:length(rem3$AGEC)){
  age = rem3$AGEC[i]
  if (age > 17 & age < 28){
    cats[i] = "18-27"
  }
  if (age > 27 & age < 38){
    cats[i] = "28-37"
  }
  if(age > 37 & age < 48){
    cats[i] = "38-47"
  }
  if(age > 47 & age < 58){
    cats[i] = "48-57"
  }
  if(age > 57 & age < 68){
    cats[i] = "58-67"
    }
  if(age > 67){
    cats[i] = "68+"
  }
}

newdf = cbind(agetrenddf, cats)
colnames(newdf) = c("Age", "Cholesterol Status", "Age Bracket")

ggplot(newdf,aes(x = `Age Bracket`, fill=`Cholesterol Status`)) + geom_histogram(stat = "count", position="dodge") + labs(y = "Frequency") + ggtitle("Status of Cholesterol by Age")


```